---
title: "<center>**Using `afrilearndata` to learn Spatial Analysis in R**</center>"
author: "<center>Wyclife Agumba Oluoch (wyclifeoluoch@gmail.com) </center>"
date: "<center>`r Sys.time()`</center>"
bibliography: 
  - bib/packages.bib
nocite: '@*'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
packages <- c("base",
              'knitr',
              'rmarkdown',
              'tidyverse',
              'afrilearndata',
              'mapsf',
              'terra',
              'raster')
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == FALSE)){
  install.packages(packages[!installed_packages])
}
lapply(packages, library, character.only = TRUE) |> invisible()
```

```{r write_bib, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::write_bib(c(
  .packages(), packages
), 'bib/packages.bib')
```

# Introduction

`afrilearndata` package has data that can be used to learn and research more about Spatial Analysis in `R` especially within the African context. Having vector (with attribute data) and raster (with pixel values), the possibilities of analytic approaches are unlimited. I will barely glimpse tip of the iceberg.

You can find the rmd file in my [GitHub](https://github.com/Wycology/afrilearn_data/blob/main/afrilearn_data.Rmd) account and more about the package [here](https://github.com/afrimapr/afrilearndata).

## Simple mapping of the countries

```{r countries}
mf_init(africountries, expandBB = rep(0, 4), theme = 'iceberg')
mf_shadow(africountries, col = 'magenta', cex = 2, add = TRUE) # just for aesthetic
mf_layout(title = 'African countries map with shadow',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n',
          'Package: mapsf ',
          packageVersion('mapsf'), '\nPackage: afrilearndata ', 
          packageVersion('afrilearndata')))
mf_map(africountries, add = TRUE)
mf_label(africountries, var = "name", col = "purple", cex = 0.9, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = TRUE)
```

## Mapping of population field as proprotional circles

```{r pop_prop}
mf_init(africountries, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African countries population estimate',
          credits = paste0('Sources: IGN, 2018\nAuthor: Wyclife Agumba Oluoch\n',
          'Package: mapsf ',
          packageVersion('mapsf'), '\nPackage: afrilearndata ',
          packageVersion('afrilearndata')))
mf_map(africountries, add = TRUE)
mf_map(x = africountries, var = 'pop_est', type = 'prop', inches = 0.25, 
       col = 'purple', leg_pos = 'topright', leg_title = 'Total population estimate'
)
```

## Mapping population in choropleth protocol

```{r pop_est_choro}
mf_init(africountries, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African countries population estimate choropleth',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = africountries, var = "gdp_md_est", type = "choro", breaks = "geom", 
       nbreaks = 5, pal = "Purples", border = "white", lwd = 0.5, leg_pos = "left",
       leg_title = "Numeric estimated gdp", add = TRUE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(africountries, col = "red")
mf_inset_off()
```

# Extracting population raster data for each country

```{r extract}
afripop2020_extract <- africountries %>% 
  mutate(pop_2020 = raster::extract(afripop2020, africountries, fun = mean, na.rm = TRUE))
```

So cool that we can extract mean values and mutate in the countries data as a new column to be used in subsequent mapping.

```{r map_extracted}
mf_init(afripop2020_extract, expandBB = rep(0, 4), theme = 'candy')
mf_layout(title = 'African countries population density 2020',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = afripop2020_extract, var = "pop_2020", type = "choro", breaks = "geom", nbreaks = 5,
  pal = "YlGnBu", border = "green", lwd = 0.5, leg_pos = "left", 
  leg_title = "African Pop. Density\n(people per km2)", add = TRUE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(afripop2020_extract, col = "red")
mf_inset_off()
```

# Selective mapping

Here I map only countries whose pop_2020 values are above 100.

```{r pop_den_above_100}
afri_pop_2020_100 <- afripop2020_extract %>% filter(pop_2020 > 100)
mf_init(afri_pop_2020_100, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African countries with over 100 people per km2 in 2020',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = afri_pop_2020_100, var = "pop_2020", type = "choro", breaks = "geom", nbreaks = 5,
  pal = "YlGnBu", border = "green", lwd = 0.5, leg_pos = "left", 
  leg_title = "African Pop. Density\n(people per km2)", add = TRUE)
mf_label(x = afri_pop_2020_100, var = "name", col = "purple", cex = 0.9, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(afri_pop_2020_100, col = "red")
mf_inset_off()
```

Here I map only countries whose population density values by 2020 fall below 20 people per square kilometre.

```{r pop_den_below_20}
afri_pop_2020_below_20 <- afripop2020_extract %>% filter(pop_2020 < 20)
mf_init(afri_pop_2020_below_20, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African countries with below 20 people per km2 in 2020',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = afri_pop_2020_below_20, var = "pop_2020", type = "choro", breaks = "geom", nbreaks = 5,
  pal = "YlOrRd", border = "green", lwd = 0.5, leg_pos = "left", 
  leg_title = "African Pop. Density\n(people per km2)", add = TRUE)
mf_label(x = afri_pop_2020_below_20, var = "name", col = "purple", cex = 0.9, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(afri_pop_2020_below_20, col = "red")
mf_inset_off()
```

Here is map of the country with maximum population density by 2020

```{r max_pop_202}
afri_pop_2020_max <- afripop2020_extract %>% filter(pop_2020 == max(pop_2020, na.rm = TRUE))
mf_init(afri_pop_2020_max, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African country with maximum people per km2 in 2020',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = afri_pop_2020_max, var = "pop_2020", type = "choro", breaks = "geom", nbreaks = 1,
  pal = "YlGnBu", border = "green", lwd = 0.5, leg_pos = "left", 
  leg_title = "African Pop. Density\n(people per km2)", add = TRUE)
mf_label(x = afri_pop_2020_max, var = "name", col = "white", cex = 2, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(afri_pop_2020_max, col = "red")
mf_inset_off()
```

Here is map of the country with least/minimum population density by 2020

```{r min_pop_2020}
afri_pop_2020_min <- afripop2020_extract %>% filter(pop_2020 == min(pop_2020, na.rm = TRUE))
mf_init(afri_pop_2020_min, expandBB = rep(0, 4), theme = 'dark')
mf_layout(title = 'African countries with minimum people per km2 in 2020',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = afri_pop_2020_min, var = "pop_2020", type = "choro", breaks = "geom", nbreaks = 1,
  col = "cyan", border = "green", lwd = 0.5, leg_pos = "left", 
  leg_title = "African Pop. Density\n(people per km2)", add = TRUE)
mf_label(x = afri_pop_2020_min, var = "name", col = "white", cex = 2, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(afri_pop_2020_min, col = "red")
mf_inset_off()
```

```{r}
land_cover <- rast(system.file("extdata","afrilandcover.grd", package="afrilearndata", mustWork = TRUE))

```


```{r raster}
# mf_init(land_cover, expandBB = rep(0, 4), theme = 'dark')
# mf_raster(land_cover, add = TRUE, bgalpha = 0.1)
```


The raster for the whole continent can be cropped and masked for a specific country extent. Here I do one for Kenya.

```{r kenya_land_cover}
kenya_land_cover <- crop(land_cover, africountries[africountries$name == 'Kenya',])
kenya_mask <- raster::mask(kenya_land_cover, vect(africountries[africountries$name == 'Kenya',]))
kenya_sf <- africountries[africountries$name == 'Kenya',]
mf_init(kenya_mask, expandBB = rep(0, 4), theme = 'dark')
mf_raster(kenya_mask, add = TRUE, bgalpha = 0.1)
mf_layout(title = 'Kenyan Land Cover',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = kenya_sf, add = TRUE, col = 'transparent', lwd = 15, border = 'purple')
mf_label(x = kenya_sf, var = "name", col = "blue", cex = 2, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(kenya_sf, col = "red")
mf_inset_off()
```

> As I said earlier, this is a mere glimpse of the iceberg tip. The scale bars in the maps could be misleading as they are in degrees and not kms. In other words, I should have projected the data as follows:

```{r reprojected}
kenya_mask_reproj <- terra::project(kenya_mask, "EPSG:21096")
kenya_sf_reproj <- st_transform(kenya_sf, "EPSG:21096")
mf_init(kenya_mask_reproj, expandBB = rep(0, 4), theme = 'dark')
mf_raster(kenya_mask_reproj, add = TRUE, bgalpha = 0.1)
mf_layout(title = 'Kenyan Land Cover',
          credits = paste0('Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n', 'Package: mapsf ',
                           packageVersion('mapsf'), '\nPackage: afrilearndata ',
                           packageVersion('afrilearndata')))
mf_map(x = kenya_sf_reproj, add = TRUE, col = 'transparent', lwd = 15, border = 'purple')
mf_label(x = kenya_sf_reproj, var = "name", col = "blue", cex = 2, font = 250, bg = "white", r = 0.4, overlap = FALSE, lines = FALSE)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(kenya_sf_reproj, col = "red")
mf_inset_off()
```

# In this case, the scale of 200km is now correct.

# References
