---
title: "Mapping Raster layer using mapsf"
author: "Wyclife Agumba Oluoch"
date: "`r Sys.time()`"
format: html
self-contained: true
theme: solar
---

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can also just click _Ctrl+Shift+K_.

```{r}
# Loading the necessary libraries for 

library(afrilearndata) 
library(mapsf)
library(raster)
library(sf)
library(terra)
```

## Versioning the libraries

It is always a good practice to have versions of packages used in check in order to avoid any possible bugs when running the same code in future.

```{r}
pkgs <- c("afrilearndata", "mapsf", "raster", "sf", "terra")

for (i in pkgs) {
  print(paste0(i, " ", "==>:", " ", packageVersion(i)))
}

```

On the other hand, to get the versions of R and Rstudio, run the following two commands in the Console:

```{r}
#| eval = FALSE
version$version.string
rstudioapi::versionInfo()$long_version
```

You can add options to executable code like this 

```{r}
#| echo: false
kenya_sf <- africountries[africountries$name == 'Kenya', ]
kenya_sp <- as_Spatial(kenya_sf)

land_cover <- rast(system.file("extdata", "afrilandcover.grd",
                               package = "afrilearndata"))

kenya_mask <- mask(crop(raster(land_cover), kenya_sp), kenya_sp)

mf_init(kenya_sf, theme = "candy")
mf_raster(rast(kenya_mask), add = TRUE)
mf_map(
  kenya_sf,
  add = TRUE,
  col = 'transparent',
  lwd = 5,
  border = 'grey40'
)
mf_scale(size = 0.0009)
mf_layout(
  title = 'Kenyan Land Cover',
  credits = paste0(
    'Sources: Andy South, 2021\nAuthor: Wyclife Agumba Oluoch\n',
    'Package: mapsf ',
    packageVersion('mapsf'),
    '\nPackage: afrilearndata ',
    packageVersion('afrilearndata')
  )
)
mf_label(
  x = kenya_sf,
  var = "name",
  col = "blue",
  cex = 1.5,
  font = 250,
  bg = "white",
  r = 0.4
)
mf_inset_on(x = 'worldmap', pos = 'topright')
mf_worldmap(kenya_sf, col = "red")
mf_inset_off()
```

The `echo: false` option disables the printing of code (only output is displayed).
